---
title: "LIFE748 - ML and data analysis"
author: "Francesco Del Carratore"
format: html
editor: visual
---

## Introduction

In this tutorial, we will analyze a simulated RNA-seq dataset. Our goal is to identify Differentially Expressed Genes (DEGs) between two conditions (Condition A and Condition B), visualize the results, and use both unsupervised and supervised machine learning techniques to classify the samples based on their gene expression profiles.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Step 0: Load the required libraries

```{r warning=FALSE, message=FALSE }
library(DESeq2) # Bioconductor
library(ggplot2) # CRAN
library(EnhancedVolcano) # Bioconductor
library(pheatmap) # Bioconductor
library(caret) # CRAN
library(MASS) # CRAN
library(e1071) # CRAN
library(groupdata2)
```

If these packages are not already available, you will need to install them either from CRAN (install.packages('package_name')), or from Bioconductor (BiocManager::install('package_name')

## Step 1: Load the Count Matrix

First, we load the raw count matrix and set up the sample metadata (phenotype data). We assume the first 25 columns belong to Condition A, and the next 25 belong to Condition B.

```{r}
counts <- read.csv("dataset_LIFE748_2.csv", row.names = 1)

colData <- data.frame(
  row.names = colnames(counts),
  Condition = c(rep('CondA',25),rep('CondB',25))
)

```

## Step 2: Normalize and Identify DEGs with DESeq2

We will use `DESeq2` to normalize the library sizes and perform statistical testing to identify which genes are differentially expressed between Condition B and Condition A.

```{r}
# 1. Create the DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = colData,
                              design = ~ Condition)

# 2. Run the DESeq2 pipeline
dds <- DESeq(dds)

# 3. Extract the results (Condition B vs Condition A)
res <- results(dds)

# 4. Convert this object into a dataframe for easier visualisation (optional)

res <- as.data.frame(res)

# 5. Extract normalized counts for downstream machine learning
vsd <- vst(dds, blind = FALSE) # Variance stabilizing transformation
normalized_counts <- assay(vsd)

```

## Step 3: Report DEGs (Table and Volcano Plot)

We define statistically significant DEGs as those with an adjusted p-value (FDR) \< 0.05 and an absolute log2 Fold Change \> 1.

```{r}
#convert to dataframe for further analysis
res <- as.data.frame(res)
#save the statistically significant DEGs into a new variable
res_sig <- res[res$padj < 0.05 & abs(res$log2FoldChange)>1, ]
```

Generate a volcano plot to visualise the results

```{r}
#| fig-width: 6
#| fig-height: 9

#volcano plot to visualise the results
EnhancedVolcano(res,
  lab = rownames(res),
  x = "log2FoldChange",
  y = "padj",
  pCutoff = 0.05,
  FCcutoff = 1,
  ylim = c(0, 8),
  pointSize = 2.0,
  labSize = 5.0
)

```

## Step 4: Isolate Differentially Expressed Genes

To reduce noise and dimensionality for our machine learning algorithms, we will subset our normalized count matrix to only include the statistically significant DEGs identified above.

In order to make sure we don't lose too much information, let's relax a bit the thresholds for significance

```{r}
#relax thresholds
res_sig <- res[res$padj < 0.1 & abs(res$log2FoldChange)>0.58, ]
#extract the count data for significant genes
sig_genes <- intersect(rownames(res_sig), rownames(normalized_counts))
#create a new dataframe to store the statistics for these significant genes
normalized_counts_sig <- normalized_counts[sig_genes, ]
```

## Step 5: Unsupervised Clustering

Let's see if the samples naturally group by their condition without explicitly telling the algorithm which sample belongs to which group. We will use Hierarchical Clustering and K-means.

## K-means

```{r}
#run elbow test
k.max <- 10
wss <- sapply(1:k.max, 
              function(k){kmeans(normalized_counts_sig[,1:2], k, nstart=5,iter.max = 15 )$tot.withinss})
plot(1:k.max, wss,
     type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")

```

```{r}
#ensure reproducibility
set.seed(123)
#find 2 clusters
k <- 2
#perform kmeans
km_samples <- kmeans(t(normalized_counts_sig), centers = k, nstart = 20)

#perform pca
pca <- prcomp(t(normalized_counts_sig), scale. = TRUE)
#save pca statistics into new dataframe
pca_df <- data.frame(PC1 = pca$x[,1],
                     PC2 = pca$x[,2],
                     Cluster = as.factor(km_samples$cluster))

#find the centers of each cluster
centers_pca <- as.data.frame(predict(pca, km_samples$centers))
centers_pca$Cluster <- as.factor(1:k)

#plot a graph with points colored the cluster predicted by kmeans
ggplot(pca_df, aes(x = PC1, y = PC2, col = Cluster)) +
  geom_point(size = 2) +
  #plot center of each cluster
  geom_point(data = centers_pca, 
             aes(x = PC1, y = PC2, shape = Cluster), 
             size = 4, color = 'black') +
  #define colors for each cluster group
  scale_color_manual(name = "Cluster Group", 
                     values = c("1" = "#E41A1C", "2" = "#377EB8")) +
  #ensure the shape and label match the data
  scale_shape_manual(name = "Centroid", 
                     values = c("1" = 17, "2" = 17), # 17 is the triangle
                     labels = c("1" = "Center 1", "2" = "Center 2")) +
  theme_minimal(base_size = 16)
```

## Hierarchical clustering

```{r}
#calculate distance matrix using euclidean method
dist_samples <- dist(t(normalized_counts_sig), method = "euclidean")

#create a ward's linkage
hc_samples <- hclust(dist_samples, method = "ward.D2")

#cut the dendrogram according to the numnber of clusters
plot(hc_samples, main = "Hierarchical Clustering of Samples", 
     xlab = "", sub = "", cex = 0.5, hang = -1)
rect.hclust(hc_samples , k=2, border = 2:6)
```

Visualise the results (HINT: an heatmap would be perfect for this).

```{r}
library(pheatmap)

#create new datafrae for producing heatmap
heatmap_df <- data.frame(
  #extract condition (ground truth) from column data
  Condition = as.factor(colData$Condition),
  KMeans = as.factor(km_samples$cluster)
)

#heatmap plotting
pheatmap(normalized_counts_sig, 
         main = "K-Means vs Hierarchical Clustering Heatmap",
         annotation_col = heatmap_df, 
         #create upper dendrogram according to previous hierarchical clustering
         cluster_cols = hc_samples,
         clustering_distance_cols = dist_samples,
         #apply z-score transformation to ensure comparable results
         scale = "row",  
         #remove left dendrogram, column and row names to decrease cluttering
         show_rownames = FALSE,
         show_colnames = FALSE,
         treeheight_row = 0,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(100))
```

## Step 6: Supervised Learning (LDA and SVM)

Next, we train classifiers to predict the condition of a sample based on its DEG expression profile. We will split the data into a training set (70%) and a testing set (30%).

```{r}
#enuure reproducibility
set.seed(122)
#create new transposed dataframe
df <- as.data.frame(t(normalized_counts_sig))
#add the condition information (ground truth)
df$Condition <- colData$Condition
df$Condition <- as.factor(df$Condition)

#split the data into training and testing sets
trainIndex <- createDataPartition(df$Condition,p=0.7, list=FALSE,times=1)
df_train <- df[trainIndex,] 
df_validate <- df[-trainIndex,]
```

Logistic regression

```{r}
#select the top 5 significant genes only for logistic regression
res_ordered <- res_sig[order(res_sig$padj), ]
top_genes <- rownames(res_ordered)[1:5]

#ensure the list of 5 genes is in the correct format
formula_string <- paste("Condition ~", paste(top_genes, collapse = " + "))
#run logistic regression
model_lr <- glm(as.formula(formula_string), data = df_train, family = "binomial")

summary(model_lr)
```

Train LDA (Linear Discriminant Analysis)

```{r}
#train the LDA model
model_LDA <- lda(Condition~., data=df_train)
model_LDA
plot(model_LDA)
```

Train SVM (Support Vector Machine)

```{r}
#train the SVM model
model_SVM <- svm(Condition~., data=df_train)
summary(model_SVM)
```

## Step 7: Compare Classifier Performance

Finally, we use confusion matrices to evaluate and compare how well LDA and SVM performed on the unseen test dataset.

```{r}
cond_levels <- levels(as.factor(df_validate$Condition))
ref_factor  <- factor(df_validate$Condition, levels = cond_levels)
pred_lr <- predict(model_lr, newdata=df_validate, type = 'response')
pred_class <- ifelse(pred_lr < 0.5, cond_levels[1], cond_levels[2])
pred_class <- factor(pred_class, levels = cond_levels)
CM_lr <- confusionMatrix(data=as.factor(pred_class), ref_factor)
CM_lr
```

```{r}
#run a prediction using the LDA model
pred_LDA <- predict(model_LDA, newdata=df_validate,type="class")
#create confusion matrix
CM_LDA = confusionMatrix(data=as.factor(pred_LDA$class), as.factor(df_validate$Condition))
CM_LDA

```

```{r}
#run prediction using SVM model
pred_SVM <- predict(model_SVM, newdata=df_validate,type="class")
#create confusion matrix
CM_SVM = confusionMatrix(data=as.factor(pred_SVM), as.factor(df_validate$Condition))
CM_SVM

```

## Step 8: Is this enough?

is this enough to compare the classifiers? What if you use a k-fold cross-validation approach? Will you get similar results with different random splits?

```{r}
#create folds for cross validation
folds <- fold(data=df, k=7)$.folds

#create empty lists to store output
CMs <- list()
models <- list()
out <- NULL

for(k in 1:7){
  #split training and testing data based on fold ID
  df_test <- df[folds==k,]
  df_train <- df[folds!=k,]
  
  #train the model
  model_lr <- glm(as.formula(formula_string), data = df_train, family = "binomial")
  
  #store model in the list
  models[[k]] <- model_lr
  
  #predict using logistic regression
  pred_lr <- predict(model_lr, newdata=df_validate, type = 'response')
  pred_class <- ifelse(pred_lr < 0.5, cond_levels[1], cond_levels[2])
  pred_class <- factor(pred_class, levels = cond_levels)
  CM <- confusionMatrix(data=as.factor(pred_class), ref_factor)
  
  #store the confusion matrix and metrics
  CMs[[k]] <- CM
  out <- rbind(out, CM$overall)
}


cv_results <- as.data.frame(out)
lr_summary_stats <- colMeans(cv_results)
print(lr_summary_stats)
```

```{r}
#create folds for cross validation
folds <- fold(data=df, k=7)$.folds

#create empty lists to store output
CMs <- list()
models <- list()
out <- NULL

for(k in 1:7){
  #split training and testing data based on fold ID
  df_test <- df[folds==k,]
  df_train <- df[folds!=k,]
  
  #train the model
  model_LDA <- lda(Condition~., data=df_train)
  
  #store model in the list
  models[[k]] <- model_LDA
  
  #predict using LDA
  pred <- predict(model_LDA, newdata=df_test)
  #create confusion matrix
  CM = confusionMatrix(data=as.factor(pred$class), as.factor(df_test$Condition))
  
  #store the confusion matrix and metrics
  CMs[[k]] <- CM
  out <- rbind(out, CM$overall)
}


cv_results <- as.data.frame(out)
lda_summary_stats <- colMeans(cv_results)
print(lda_summary_stats)

```

```{r}
#create folds for cross validation
folds <- fold(data=df, k=7)$.folds

#create empty lists to store output
CMs <- list()
models <- list()
out <- NULL

for(k in 1:7){
  #split training and testing data based on fold ID
  df_test <- df[folds==k,]
  df_train <- df[folds!=k,]
  
  #train the model
  model_SVM <- svm(Condition~., data=df_train)
  
  #store model in the list
  models[[k]] <- model_SVM
  
  #predict using LDA
  pred <- predict(model_SVM, newdata=df_test)
  #create confusion matrix
  CM = confusionMatrix(data=as.factor(pred), as.factor(df_test$Condition))
  
  #store the confusion matrix and metrics
  CMs[[k]] <- CM
  out <- rbind(out, CM$overall)
}

cv_results <- as.data.frame(out)
svm_summary_stats <- colMeans(cv_results)
print(svm_summary_stats)

```

## Interpret Biologically Important Markers

```{r}
#retrieve centers from k means testing
centers <- km_samples$centers 

#calculate the difference of geneexpression between the centers of cluster 1 and 2
gene_importance_km <- abs(centers[1, ] - centers[2, ])

#find the top 2 genes that contribute most to the clustering
top_km_markers <- sort(gene_importance_km, decreasing = TRUE)[1:2]
print(top_km_markers)

```

```{r}
#calculate the weight vector by multiplying coefficients by the support vectors
w <- t(model_SVM$coefs) %*% model_SVM$SV

#find the genes with the highest absolute weights
svm_weights <- abs(as.vector(w))
#create a list of gene names
names(svm_weights) <- colnames(df_train)[-ncol(df_train)] 
#match the svm weights to each gene and extract the top 10 genes with the highest weights
top_svm_markers <- sort(svm_weights, decreasing = TRUE)[1:10]
print(top_svm_markers)

#extract the scaling coefficients from the LDA model
lda_importance <- abs(model_LDA$scaling[,1])

#find top 10 genes that contribute most to the LDA linear discriminant
top_lda_markers <- sort(lda_importance, decreasing = TRUE)[1:10]
print(top_lda_markers)

#find common genes for the LDA and SVM method
supervised_markers <- intersect(names(top_svm_markers), names(top_lda_markers))
print(supervised_markers)
```

## Data preparation for part 3

```{r}
#create a dataframe with the gene IDs and fold changes
res_sig_ordered <- res_sig[order(abs(res_sig$log2FoldChange), decreasing = TRUE), ]
export_df <- data.frame(
  GeneID = rownames(res_sig_ordered),
  log2FoldChange = res_sig_ordered$log2FoldChange,
  padj = res_sig_ordered$padj
)

#write the data into a new tsv file
write.table(export_df, 
            file = "significant_genes_with_FC.txt", 
            sep = "\t",
            row.names = FALSE,
            col.names = TRUE,
            quote = FALSE)
```
